---
title: "Example Manuscript Template for a Data Analysis Project"
author: ""
format:
  docx:
    toc: false
    number-sections: true
    highlight-style: github
bibliography: ../../assets/dataanalysis-references.bib
csl: ../../assets/american-journal-of-epidemiology.csl
---

The structure below is one possible setup for a manuscript stemming from a data analysis project. It loosely follows the structure of a standard scientific manuscript. Adjust as needed. You don't need to have exactly these sections, but the content covering those sections should be addressed.

This uses [MS Word as output format](https://quarto.org/docs/output-formats/ms-word.html). See [the Quarto documentation](https://quarto.org/) for instructions on how to use other formats.


```{r, echo=FALSE, message=FALSE}
# load a few R packages
library(here)
library(knitr)
```


**Authors**  

* Esther Palmer$^{1}$ (ORCID: ) 
* Dawson Deal$^{1}$
* Kaya Palukaitis$^{1}$
* Elisabeth Drake$^{2}$ 
* Erin Lipp$^{2}$ 
* Nikki Shariat$^{1, \land}$ 

**Author affiliations**  

1. Departments of Microbiology and Population Health, University of Georgia, Athens, GA, USA.
2. Environmental something? will fix later, University of Georgia, Athens, GA, USA.


$*$ These authors contributed equally to this work.

$\land$ Corresponding author: some@email.com

$\dagger$ Disclaimer: The opinions expressed in this article are the
author's own and don't reflect their employer.

I (Esther) will be the only one doing modeling on this project, however I do want to acknowledge the undergrads who did a majority of the actual data collection (Dawson and Kaya) because they are fantastic, as well as our collaborators in other labs who helped with some of the other data generation.

{{< pagebreak >}}



# Summary/Abstract


# Introduction 

## General Background Information
_Salmonella enterica_ is an enteric pathogen that alternates between growth in a host and survival in environmental reservoirs, such as in freshwater, soil, and dust (Gorski et al., 2011; Li et al., 2015; Pal et al., 2022). _Salmonella_ has been repeatedly isolated from surface water across the United States with prevalence ranging from 4-100% (Deaven et al., 2021; Gorski et al., 2011; Havelaar et al., 2017; Maurer et al., 2015; Smith et al., 2025). Approximately 52% of the irrigation water in the US comes from surface water (Dieter et al., 2018), and contaminated surface water has been shown to be a risk for produce contamination (FDA, 2025, 2024; Greene et al., 2008). _Salmonella_ is an highly diverse organism with over 2,600 different serovars (Grimont, n.d.). However, not all serovars pose equal food safety risks. Some are considered host-restricted, able to infect only one host. For example, serovar Typhi is a human host-restricted serovar responsible for typhoid fever. Others are considered host-adapted, able to best infect one host and sometimes others. Serovar Dublin, which is primarily associated with cattle but also sometimes responsible for human illness, is one instance of this. And some, such as serovar Typhimurium, are considered host-promiscuous, able to infect a wide variety of hosts (Uzzau et al., 2000). As such it is important to study _Salmonella_ population dynamics on a serovar level. While _Salmonella_ prevalence in surface water is well characterized, not much is known about how complex serovar populations are in surface water, nor how frequently these populations turn over, if at all. Most studies either sample a large number of sites at one timepoint, or sample a smaller number of sites a few times, often with weeks, months, or years in between sampling. This does not provide sufficient temporal resolution to assess how frequently _Salmonella_ populations change and what factors can affect these changes. It should also be noted few of these studies capture the full variety of salmonellae in the environment as most rely on the characterization of single isolates, which can miss serovars that are present at lower relative frequencies.

Where there have not been many attempts to model for _Salmonella_ complexity, likely due to the paucity of data on _Salmonella_ populations in water, there have been many attempts to model _Salmonella_ prevalence in water (Gorski et al., 2011; Haley et al., 2009; Smith et al., 2025). These papers have often been contradictory, with some finding certain factors such as rainfall or temperature to be important (Deaven et al., 2021; D. Weller et al., 2020) and others finding these factors unimportant or even negatively correlated (McEgan et al., 2013; Strawn et al., 2013). One major systemic review on _Salmonella_ prevalence in water found that sampling methodology among other things can affect comparability between studies (D. L. Weller et al., 2024). Variations in Salmonella detection due to sampling methodology could easily transfer to variations in model outcomes. In the end, much of the evidence points towards modeling only being effective for the given system.

## Description of data and data source

Our lab went sampling for _Salmonella_ at an urban creek every weekday for 6 months, and then once a week for another 9 months. _Salmonella_ prevalence during daily sampling was extremely high with 97% (n=127/131) of samples positive. CRISPR-SeroSeq was performed on all positive samples, and generated serovar profiles for 106 samples. There were 26 serovars identified in total.
Environmental data was collected at time of sampling, as well as weather data from the nearest UGA weather station (Georgia Weather - Automated Environmental Monitoring Network Page, n.d.), for a total of 20 variables total.


## Questions/Hypotheses to be addressed

The objective of this project was to use CRISPR SeroSeq, a deep serotyping technique, to assess the following: 
Evaluate changes to _Salmonella_ serovar populations surface water over time, and,
Determine what environmental and/or weather factors influence _Salmonella_ population complexity.

# Methods 



## Data acquisition
The study location sits on a suburban creek surrounded by riparian zones. This site was chosen due to the observation in a previous study that the serovars present were markedly different to those in proximal sites up and down stream (Smith et al., 2025). Sampling was conducted every weekday for 6 months, for a total of 162 sample collections. A carboy was used to collect 10L of creek water that was transported to the lab and immediately filtered through a modified Moore swab. Buffered peptone water, (BPW) was added to the swabs and incubated at 42°C for 24 hours. Three different selective enrichment media Tetrathionate (TT), Rappaport-Vassiliadis (RV), and modified semisolid Rappaport-Vassiliadis (MSRV) were inoculated with the BPW culture and incubated at 37°C overnight. These different media were chosen to maximize _Salmonella_ recovery, specifically to improve the isolation of different _Salmonella_ serovars. The cultures were then plated onto xylose lysine tergitol-4 agar (XLT4) and brilliant green agar (BGS). DNA was isolated from 1mL of the TT and RV cultures that were _Salmonella_ positive and used as a template for CRISPR-SeroSeq analysis.
Deep serotyping was performed as described (Smith et al., 2025). Briefly, at the time of culture, 1 mL each of BPW preenriched culture and RV and TT selectively enriched cultures were transferred to a microcentrifuge tube and centrifuged at 14,000 rpm for 3 minutes. After removing the supernatant, the pellets were stored at -20℃. Total genomic DNA was isolated from TT and RV pellets corresponding to _Salmonella_-positive cultures using Promega Genome Wizard kit (Madison, WI) according to the manufacturer’s instructions and resuspended in 200 µL of molecular-grade water and stored at -20℃. If neither RV nor TT was positive but MSRV was, then 1mL of the BPW was used for DNA isolation instead. PCR was performed targeting the conserved direct repeats of the CRISPR array within each sample, primers contained index sequences to facilitate multiplexed high throughput sequencing. PCR products were purified using the AMPure system (Beckman Coulter, Indianapolis, IN). Pooled libraries were sequenced using the Illumina NextSeq 550 platform. Resultant sequences were assigned to known CRISPR arrays to determine the number and identify of all serovars within a sample. A threshold of 1000 reads in a sample was applied. Serovars were only called if they had 0.5% of reads or more assigned to them. DESeq2 was applied to normalize reads across all samples (Love et al., 2014). When possible, RV and TT libraries from the same sample were combined.
It should be noted that several serovars within this study are polyphyletic, meaning that there are multiple lineages of the same serovar with different genetic backgrounds but the same O and H antigens. DST is able to distinguish between these different lineages as they often have different CRISPR arrays. As such, these lineages are denoted using roman numerals. 



## Data import and cleaning
_Write code that reads in the file and cleans it so it's ready for analysis. Since this will be fairly long code for most datasets, it might be a good idea to have it in one or several R scripts. If that is the case, explain here briefly what kind of cleaning/processing you do, and provide more details and well documented code somewhere (e.g. as supplement in a paper). All materials, including files that contain code, should be commented well so everyone can follow along._

```{r}
library(readxl) #for loading Excel files
library(dplyr) #for data processing/cleaning
library(tidyr) #for data processing/cleaning
library(skimr) #for nice visualization of data 
library(here) #to set paths
library(ggplot2) #for figures
```
Loads packages

```{r}
#path to data
enviro_data_location <- here::here("data","raw-data","Environmental_Data_Final.xlsx")
CSS_data_location <- here::here("data", "raw-data", "Daily_Sampling_Normalized_CSS_6mo.xlsx")
#load data.
#there are 2 files, one containing the environmental/weather data, and one containing CSS data
enviro_rawdata <- readxl::read_excel(enviro_data_location, sheet = "Sheet2")
#I am going to use the dataset that excludes weekends for this preliminary look, Sheet1 contains the data that has weekends
CSS_rawdata <- readxl::read_excel(CSS_data_location)
#This will read in the CSS data
```
This code chunk loads my raw data. I have two files, one with completed CSS analysis, and another with environmental and weather variables. The environmental and weather variable excel sheet has 2 sheets, on sheet1 there is all the data including the weather data for the weekends. On sheet2 the weekend data was removed and only data for the day of sampling was left.

```{r}
dplyr::glimpse(enviro_rawdata)
#looking at the raw environmental data
summary(enviro_rawdata)
```

This provides a look at all the environmental data. I can see several variables have been listed as characters when they need to be numeric, and there are several NAs so I will have some cleaning to do in the future. I also don't trust DO I think the instrument was acting up so I kinda want to drop the whole variable. Day is also being treated like a numeric variable rather than an ordinal (I think ordinal? it has order) variable. Although maybe it is quantitative since you can do things like addition to it?

## Statistical analysis

I plan to train my models on the data collected from this site daily for 6 months, then test it on data collected from the same site weekly for another 9? months.

{{< pagebreak >}}


# Results

## Exploratory/Descriptive analysis

_

## Basic statistical analysis



## Full analysis



# Discussion

## Summary and Interpretation
_Summarize what you did, what you found and what it means._

## Strengths and Limitations
_Discuss what you perceive as strengths and limitations of your analysis._

## Conclusions
_What are the main take-home messages?_

_Include citations in your Qmd file using BibTeX, the list of references will automatically be placed at the end_



# References

Love, M. I., Huber, W., & Anders, S. (2014). Moderated estimation of fold change and dispersion for RNA-seq data with DESeq2. Genome Biology, 15(12), 550. https://doi.org/10.1186/s13059-014-0550-8

Smith, J. C., Siceloff, A. T., Shirazi, S. M., Bell, R. L., & Shariat, N. W. (2025). A two-year study of Salmonella in four natural watersheds highlights the need for increased environmental Salmonella surveillance to close the One Health loop. Applied and Environmental Microbiology, 91(12), e01770-25. https://doi.org/10.1128/aem.01770-25

